---
title: "DATA_316_Cyclic_Presentation"
author: "Riley,Emily,Iyanu,Bryan"
date: "2023-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Read in the Sunrise Dataset

```{r}
suntimes <- read.csv("https://raw.githubusercontent.com/gmtanner-cord/DATA316/main/Data/MoorheadSunriseSunset.csv",stringsAsFactors = TRUE)
suntimes$Date <- as.Date(suntimes$Date,format="%m/%d/%Y") #https://www.statology.org/convert-strings-to-dates-in-r/
suntimes <- suntimes %>% mutate(Day=as.POSIXlt(Date)$yday+1) #https://rc2e.com/stringsanddates
suntimes$Day <- as.integer(suntimes$Day)
```

# Configure the initial Cyclic statistical model
```{r}

sunrise_model <- nls(Sunrise.CST ~ a * sin((b * Day) - c) + d,suntimes,start=list(a=mean((max(suntimes$Sunrise.CST)-mean(suntimes$Sunrise.CST))+(min(suntimes$Sunrise.CST)+mean(suntimes$Sunrise.CST))), b=(365/(2*pi)), c=182.5, d=mean(suntimes$Sunrise.CST)))
sunrise_model

mean((max(suntimes$Sunrise.CST)-mean(suntimes$Sunrise.CST))+(min(suntimes$Sunrise.CST)+mean(suntimes$Sunrise.CST)))
mean(suntimes$Sunrise.CST)

suntimes$sunrise_predicted <- predict(sunrise_model)

ggplot(suntimes,mapping=aes(x=Day,y=Sunrise.CST)) +
  geom_point() +
  geom_line(mapping=aes(x=Day,y=sunrise_predicted))

```

# Influence of the Coefficients

For $$ y = a \sin(bt - c) + d $$ , a is the amplitude, b is the period times 2$$pi$$, c is the phase shift, and d is the vertical shift. (source: http://jwilson.coe.uga.edu/EMAT6680Fa09/Gonterman/Gonterman1/Gonterman1.html)
Thus for selecting values for our model:


- Let a = mean((max(Sunrise.CST)-mean(Sunrise.CST))+(min(Sunrise.CST)+mean(Sunrise.CST))), or in other words, the average of the distance from the maximum to the mean and from the minimum to the mean
- Let b = 365/2pi, or a full range of days in a year since that is our period
- Let c = 182.5, or half a year to start the sine graph off at the maximum of the wave
- Let d = mean(Sunrise.CST), or the middle of the data
