---
title: "DATA_316_Cyclic_Presentation"
author: "Riley,Emily,Iyanu,Bryan"
date: "2023-01-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Read in the Sunrise Dataset

```{r}
suntimes <- read.csv("https://raw.githubusercontent.com/gmtanner-cord/DATA316/main/Data/MoorheadSunriseSunset.csv",stringsAsFactors = TRUE)
suntimes$Date <- as.Date(suntimes$Date,format="%m/%d/%Y") #https://www.statology.org/convert-strings-to-dates-in-r/
suntimes <- suntimes %>% mutate(Day=as.POSIXlt(Date)$yday+1) #https://rc2e.com/stringsanddates
suntimes$Day <- as.integer(suntimes$Day)
```

# Configure the initial Cyclic statistical model
```{r}

sunrise_model <- nls(Sunrise.CST ~ a * sin((b * Day) - c) + d,suntimes,start=list(a=mean((max(suntimes$Sunrise.CST)-mean(suntimes$Sunrise.CST))+(min(suntimes$Sunrise.CST)+mean(suntimes$Sunrise.CST))), b=(365/(2*pi)), c=182.5, d=mean(suntimes$Sunrise.CST)))
sunrise_model

mean((max(suntimes$Sunrise.CST)-mean(suntimes$Sunrise.CST))+(min(suntimes$Sunrise.CST)+mean(suntimes$Sunrise.CST)))
mean(suntimes$Sunrise.CST)
min(suntimes$Sunrise.CST)

suntimes$sunrise_predicted <- predict(sunrise_model)

ggplot(suntimes,mapping=aes(x=Day,y=Sunrise.CST)) +
  geom_point() +
  geom_line(mapping=aes(x=Day,y=sunrise_predicted))

```

# Influence of the Coefficients

For $$ y = a \sin(bt - c) + d $$ , a is the amplitude, b is the period times 2$$\pi$$, c is the phase shift, and d is the vertical shift. (source: http://jwilson.coe.uga.edu/EMAT6680Fa09/Gonterman/Gonterman1/Gonterman1.html)
Thus for selecting values for our model (used desmos for fitting to our dataset: https://www.desmos.com/calculator/ueu84lwtkn


- Let a = 190
- Let b = 1/(2pi*9.24)
- Let c = 4.7
- Let d = 622

Why these values?
622 = mean(max(Sunrise.CST),min(Sunrise.CST)) (d will be the point at which our sine wave is halfway between max and min)
190 = max(Sunrise.CST)-622 (the range that needs to be on either side of d, so that our maximum point is equal to the maximum point and our minimum point is equal to the minimum point in our dataset)
b & c were found using the sliders on Desmos to match the general shape that we want the curve to fit (365=period*(3/4))


# Second Attempt at Model
```{r}

sunrise_model <- nls(Sunrise.CST ~ a * sin((b * Day) - c) + d,suntimes,start=list(a=190,b=1/(2*pi*9.24),c=4.7, d=622))
sunrise_model

mean((max(suntimes$Sunrise.CST)-mean(suntimes$Sunrise.CST))+(min(suntimes$Sunrise.CST)+mean(suntimes$Sunrise.CST)))
mean(suntimes$Sunrise.CST)

suntimes$sunrise_predicted <- predict(sunrise_model)

ggplot(suntimes,mapping=aes(x=Day,y=Sunrise.CST)) +
  geom_point() +
  geom_line(mapping=aes(x=Day,y=sunrise_predicted))


```