---
title: "Homework_3_Math_Probability_Model"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

##To plot a rise and fall graph, we had  to change the date in m/dd/yyyy to day of the year for easy plotting of the Graph.
##Aslo, we change the sunrise and sunset cst to hours and mins.
```{r}
MoorheadSunriseSunset <- read_csv("https://raw.githubusercontent.com/gmtanner-cord/DATA316/main/Data/MoorheadSunriseSunset.csv") %>%
  mutate(across(Date,parse_date,format="%m/%d/%Y")) %>%
mutate(SunriseHR =floor(`Sunrise CST`/100))  %>%
  mutate(SunriseMin = `Sunrise CST`-100*SunriseHR) %>%
mutate(SunsetHR =floor(`Sunset CST`/100))  %>% 
  mutate(SunsetMin = `Sunset CST`-100*SunsetHR) %>%
  mutate(SunriseTime= SunriseHR +SunriseMin/60) %>%
 mutate(SunsetTime= SunsetHR +SunsetMin/60) %>%
  mutate(DOY=yday(Date))
str(MoorheadSunriseSunset)
MoorheadSunriseSunset
```

##sunrisetime and sunsettime is the time in hours plus mintue divided by 60.
##plotting a graph of sunrisetime against Date.
```{r}
ggplot(MoorheadSunriseSunset,aes(x=Date))+
  labs(title = "Cyclic",
       x = "Date",
       y ="Sunrise")+
  geom_line(aes(y=SunriseTime))
```



$$ y = a \sin(bt - c) + d $$
`data= MoorheadSunriseSunset`
`start=list(a= amplitide b=2π/b, c=horizontal shift , d=vertical shift)`
##formular: `Sunrise_CST ~ d+a sin(b*DOY-c))`


```{r}
mod.Sunrise = nls(SunriseTime ~ d+a*sin(b*DOY-c), data=MoorheadSunriseSunset,control=list(warnOnly=TRUE),start=list(a=1.75, d=6.5, b=2*3.14/365, c= -3.14/2))
mod.Sunrise

mod.Sunset = nls(SunsetTime ~ d+a*sin(b*DOY-c), data=MoorheadSunriseSunset,control=list(warnOnly=TRUE),start=list(a=1.75, d=6.5, b=2*3.14/365, c= -3.14/2))
mod.Sunset

##MoorheadSunriseSunset$predictedQ = predict(mod.Sunrise)
##MoorheadSunriseSunset$predictedQ = predict(mod.Sunset)
```
##plot of sunsettime of both the actual and predicted value.
##predicted is in blue.

```{r}
ggplot(MoorheadSunriseSunset,aes(x=Date))+
  labs(title = "Cyclic",
       x = "Date",
       y ="Sunset")+
  geom_line(aes(y=SunsetTime))+
  geom_line(aes(y=predict(mod.Sunset)), color= "blue")
```

```{r}
coef(mod.Sunset)
summary(mod.Sunset)
predict(mod.Sunset)
residuals(mod.Sunset)

```

##plot of sunrisetime of both the actual and predicted value.
##predicted is in green.
```{r}
ggplot(MoorheadSunriseSunset,aes(x=Date))+
  labs(title = "Cyclic",
       x = "Date",
       y ="Sunrise")+
  geom_line(aes(y=SunriseTime))+
  geom_line(aes(y=predict(mod.Sunrise)), color= "green")
```


```{r}
coef(mod.Sunrise)
summary(mod.Sunrise)
predict(mod.Sunrise)
residuals(mod.Sunrise)
```


```{r}
##To show pattern in rises and falls that are not of fixed period.

##During signal processing

##They are used in sciences ( to describe waves pattern)

```


